---
title: "Estimating mammal density using the cam_*() workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating mammal density using the cam_*() workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval    = FALSE,
  echo    = TRUE,
  comment = "#>"
)
```

## Overview

The `cam_*()` family of functions can be used to implement a **time-in-front-of-camera (TIFC)** pipeline
for estimating wildlife density at individual camera locations^[The TIFC method is described in [Becker et al. 2022](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.4005)]. The method quantifies how much total time each species spends in the camera's field of view, then converts that to an area-standardised density estimate (individuals/km²) using the camera's effective detection distance (EDD) and the number of days it was operational.

These functions are designed to work with the reports (i.e., output) from [WildTrax](www.wildtrax.ca), which can be read into the user's R environment either as a csv file or using the data download functions available in the [wildrtrax](https://abbiodiversity.github.io/wildrtrax/) R package.

---

## Prerequisites

```{r libraries}
# Load packages
library(wildrtrax)
library(tidyverse)
library(sciCentRverse)
```

The `cam_*()` functions depend on several internal lookup tables stored in the
package's `sysdata.rda`:

| Object | Contents |
|---|---|
| `native_sp` | Character vector of recognized wildlife species names from WildTrax |
| `tbi` | Per-species average time-between-photos (seconds) |
| `dist_groups` | Mapping of species to EDD distance groups |
| `edd` | EDD lookup by dist_group × vegetation category × season × model × height |

These are loaded automatically with the package and do not need to be supplied by the user.

---

## Step 1: Download data from WildTrax

The functions in this workflow expect the output from WildTrax. Data can be downloaded using the `wildrtrax` package. The example below downloads the **main** and **image** reports from the ABMI OSM 2021 project. 

```{r download}
# Authenticate into WildTrax
wt_auth()

# Find projects of interest
osm_proj <- wt_get_projects(sensor = "CAM") |>
  filter(project == "ABMI OSM 2021")

# Download reports for all projects
ls_osm_reports <- map(
  .x = osm_proj$project_id,
  .f = ~ wt_download_report(
    project_id = .x,
    sensor_id  = "CAM",
    reports    = c("main", "image_report")
  )
)
```

Extract and tidy image and main reports separately:

```{r tidy-reports}
# Image report: one row per image
df_image_reports <- ls_osm_reports |>
  list_flatten() |>
  (\(x) x[str_detect(names(x), "image_report\\.csv$")])() |>
  list_rbind() |>
  left_join(osm_proj, by = "project_id") |>
  select(project, project_id, location, location_id,
         image_id, image_date_time, image_fov,
         image_trigger_mode, equipment_model)

# Main report: one row per image × species tag
df_main_reports <- ls_osm_reports |>
  list_flatten() |>
  (\(x) x[str_detect(names(x), "main_report\\.csv$")])() |>
  list_rbind() |>
  left_join(osm_proj, by = "project_id") |>
  select(project, project_id, location, location_id,
         image_id, image_date_time, species_common_name,
         individual_count, age_class, sex_class, tag_id)
```

---

## Step 2: Camera operating days

`cam_get_op_days()` builds a per-day operational calendar for each camera. A
day is counted as **operational** if at least one in-range image exists (i.e.
`image_fov != "OOR"`). Images triggered by `"CodeLoc Not Entered"` are
excluded entirely.

```{r op-days}
df_days <- df_image_reports |>
  cam_get_op_days(
    grouping   = c("project_id", "project", "location_id", "location"),
    missing_as = TRUE,
    span       = "data"
  )
```

The result is a long tibble with one row per camera per day, with an `operating`
logical column.

### Summarising by season

`cam_summarise_op_by_season()` aggregates the daily calendar into seasonal
operating-day totals. Seasons are defined by **Julian day start cutoffs** and
are applied consistently across all subsequent pipeline steps.

```{r op-season}
# Standard ABMI season definitions
seasons <- c(spring = 99L,   # ~April 9
             summer = 143L,  # ~May 23
             winter = 288L)  # ~October 15

df_days_summary <- df_days |>
  cam_summarise_op_by_season(
    seasons  = seasons,
    by_year  = FALSE,
    wide     = TRUE   # one column per season + total_days
  )
```

The wide output has one row per camera with columns `spring`, `summer`,
`winter`, and `total_days` (all counts of operational days).

---

## Step 3: Consolidate species tags

WildTrax stores tags at the **image × individual** level: if three deer are
visible in one image, there may be three rows. `cam_consolidate_tags()` collapses
these into one row per **image × species**, summing counts and concatenating
age/sex classes.

```{r consolidate}
df_main_reports_cons <- cam_consolidate_tags(df_main_reports)
```

::: {.callout-note}
The function also detects and warns about likely duplicate tag entries (identical
rows except `tag_id`) and drops unknown `species_common_name` labels with a
warning — both indicate upstream data quality issues worth investigating.
:::

---

## Step 4: Time in front of camera

### Parse into series

`cam_calc_time_by_series()` groups consecutive images of the same species at
the same camera into **series** using a time-gap threshold. Images more than
`split_gap_secs` seconds apart (default 120 s) start a new series. Each
series is then assigned a duration using the time-in-front-of-camera (TIFC)
method: bookend images receive half the gap to the next/previous image plus
half a time-between-photos (tbp) interval; interior images receive half the gap
to both neighbours.

```{r series}
df_series <- df_main_reports_cons |>
  cam_calc_time_by_series(split_gap_secs = 120)
```

#### Optional: NONE-bridged series splitting

A subtlety arises when an animal briefly leaves the field of view (triggering a
`NONE` image) and returns within the gap threshold. This pair of detections
would be joined into a single series, over-estimating time spent. 
`cam_obtain_n_gap_class()` identifies these boundaries, and the result can be
passed to `cam_calc_time_by_series()` via `n_gap_df` to force a split:

```{r n-gap}
n_gaps   <- cam_obtain_n_gap_class(df_main_reports_cons)

df_series <- df_main_reports_cons |>
  cam_calc_time_by_series(
    split_gap_secs = 120,
    n_gap_df       = n_gaps
  )
```

The output is one row per series with columns `project`, `location`,
`species_common_name`, `series_num`, `n_images`, `series_total_time` (seconds),
`series_start`, and `series_end`.

### Sum to seasonal totals

`cam_sum_total_time()` rolls up series-level times to seasonal totals per
camera × species, zero-filling locations with no detections for any species in
`species_universe`.

```{r total-time}
sp_uni <- c("White-tailed Deer", "Black Bear", "Moose",
            "Woodland Caribou", "Fisher", "Marten",
            "Coyote", "Snowshoe Hare", "Canada Lynx")

df_dur <- df_series |>
  cam_sum_total_time(
    season_cutoffs   = seasons,
    op_days_df       = df_days_summary,
    species_universe = sp_uni
  )
```

The output has one row per camera × species × season with `total_duration`
(seconds of TIFC) and `total_season_days` (operational days in that season).

---

## Step 5: Calculate density

### Camera model lookup

EDD values vary by camera model. `cam_extract_model_lookup()` derives a
`model` label (`"hf2"` or `"pc900"`) for each camera from the `equipment_model`
field in the image report.

```{r model}
df_model <- df_image_reports |>
  cam_extract_model_lookup(
    keys      = c("project", "project_id", "location", "location_id"),
    model_col = "equipment_model"
  )
```

### Density calculation

`cam_calc_density_by_loc()` joins EDD values from the internal lookup table
to each camera × species × season row and applies the TIFC density formula.
EDD lookups are keyed on species distance group, vegetation category
(`overall_category`), season, camera model, and mounting height.

```{r density}
df_density <- df_dur |>
  left_join(df_model, by = c("project", "location")) |>
  mutate(height = "high") |>              # 1 m mounting height
  filter(total_season_days >= 30) |>      # minimum effort threshold
  cam_calc_density_by_loc(
    edd_category_df     = df_osm_edd,     # vegetation category per location
    cam_fov_angle       = 40,             # degrees
    format              = "long",
    aggregate           = TRUE,           # weighted mean across seasons
    use_global_edd      = TRUE,           # fall back to pooled EDD if needed
    annotate_edd_source = TRUE            # label EDD provenance
  )
```

#### EDD source annotation

When `annotate_edd_source = TRUE`, the output includes an `edd_source` column
with three possible values:

| Value | Meaning |
|---|---|
| `"observed"` | EDD from directly measured detections (`n > 0`) |
| `"prefilled"` | EDD borrowed from a related model/height combination within the same vegetation category |
| `"pooled"` | No matching vegetation category; EDD pooled across categories at runtime |

When `aggregate = TRUE`, the most conservative level across seasons is
reported (pooled > prefilled > observed).

#### Seasonal exclusions

By default, Black Bear detections in winter are excluded from the seasonal
weighted mean before aggregating (bears are hibernating and zero detections
do not reflect low density). This is controlled by `agg_exclude_species` and
`agg_exclude_season`:

```{r exclude}
# Default: exclude Bear × winter
cam_calc_density_by_loc(
  ...,
  agg_exclude_species = "Bear",
  agg_exclude_season  = "winter"
)

# To include all seasons:
cam_calc_density_by_loc(
  ...,
  agg_exclude_species = NULL
)
```

---

## Full pipeline at a glance

```{r full-pipeline}
# 1. Operating days
df_days <- df_image_reports |>
  cam_get_op_days(grouping = c("project_id", "project", "location_id", "location"),
                  span = "data", missing_as = TRUE)

df_days_summary <- df_days |>
  cam_summarise_op_by_season(seasons = c(spring = 99L, summer = 143L, winter = 288L),
                             wide = TRUE)

# 2. Consolidate tags
df_main_reports_cons <- cam_consolidate_tags(df_main_reports)

# 3. Time in front of camera
df_series <- df_main_reports_cons |>
  cam_calc_time_by_series(split_gap_secs = 120,
                          n_gap_df = cam_obtain_n_gap_class(df_main_reports_cons))

df_dur <- df_series |>
  cam_sum_total_time(season_cutoffs   = c(spring = 99L, summer = 143L, winter = 288L),
                     op_days_df       = df_days_summary,
                     species_universe = sp_uni)

# 4. Density
df_model <- df_image_reports |>
  cam_extract_model_lookup(keys = c("project", "location"))

df_density <- df_dur |>
  left_join(df_model, by = c("project", "location")) |>
  mutate(height = "high") |>
  filter(total_season_days >= 30) |>
  cam_calc_density_by_loc(edd_category_df     = df_osm_edd,
                          cam_fov_angle       = 40,
                          aggregate           = TRUE,
                          use_global_edd      = TRUE,
                          annotate_edd_source = TRUE)
```
