---
title: "Estimating mammal density using the cam_*() workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating mammal density using the cam_*() workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval    = TRUE,
  echo    = TRUE,
  comment = "#>",
  message = FALSE,
  cache   = FALSE
)

# Add CSS for horizontal scrolling in wide outputs
default_output_hook <- knitr::knit_hooks$get("output")
knitr::knit_hooks$set(output = function(x, options) {
  out <- default_output_hook(x, options)
  if (!is.null(options$scroll) && options$scroll) {
    paste0('<div class="scroll-output">', out, "</div>")
  } else {
    out
  }
})
```

```{css, echo=FALSE}
.scroll-output pre {
  white-space: pre !important;
  overflow-x: auto !important;
}
```

## Overview

The `cam_*()` family of functions can be used to implement a **time-in-front-of-camera (TIFC)** pipeline for estimating wildlife density at individual camera locations^[The TIFC method is described in [Becker et al. 2022](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.4005)]. The method calculates how much total time each species spends in the camera's field of view, then converts that to an area-standardised density estimate (individuals/km²) using the camera's effective detection distance (EDD) and the number of days it was operational.

These functions are designed to work with the reports (i.e., output) from [WildTrax](www.wildtrax.ca), which can be read into the user's R environment either as a csv file or using the data download functions available in the [wildrtrax](https://abbiodiversity.github.io/wildrtrax/) R package.

---

## Prerequisites

```{r libraries}
# Load packages
library(wildrtrax)
library(tidyverse)
library(sciCentRverse)
```

The `cam_*()` functions depend on several internal lookup tables stored in the
package's `sysdata.rda`:

| Object | Contents |
|---|---|
| `native_sp` | Character vector of recognized wildlife species names from WildTrax |
| `tbi` | Per-species average time-between-images (seconds) |
| `dist_groups` | Mapping of species to EDD distance groups |
| `edd` | EDD lookup by dist_group × vegetation category × season × model × height |
| `gap_groups` | Grouping of time-gap classes for probabilistic gap adjustment |
| `leave_prob_pred` | Predicted probability of departure by gap_group and species_group |

These are loaded automatically with the package and do not need to be supplied by the user.

---

## Step 1: Download data from WildTrax

The functions in this workflow expect the output from WildTrax. Data can be downloaded using the `wildrtrax` package. The example below downloads the **main** and **image** reports from the ABMI's [Biodiversity Trajectories 2023](https://portal.wildtrax.ca/cam/2025) project using the `wt_download_report()` function from wildrtrax.

```{r download, cache=TRUE, cache.lazy=FALSE}
# Authenticate into WildTrax
wt_auth()

# Find project of interest
bdt_proj <- wt_get_projects(sensor = "CAM") |>
  filter(project == "Biodiversity Trajectories 2023") |>
  select(project, project_id)

bdt_proj_id <- bdt_proj$project_id

# Download reports for all projects
ls_bdt_reports <- wt_download_report(
  project_id = bdt_proj_id,
  sensor_id = "CAM",
  reports = c("main", "image_report")
)
```

Extract and tidy image and main reports separately:

```{r tidy-reports}
# Image report: one row per image
df_image_report <- ls_bdt_reports |>
  (\(x) x[str_detect(names(x), "image_report\\.csv$")])() |>
  list_rbind() |>
  left_join(bdt_proj, by = "project_id")

# Main report: one row per image × species tag
df_main_report <- ls_bdt_reports |>
  (\(x) x[str_detect(names(x), "main_report\\.csv$")])() |>
  list_rbind() |>
  left_join(bdt_proj, by = "project_id")
```

---

## Step 2: Camera operating days

`cam_get_op_days()` builds a per-day operational calendar for each camera from the image report dataframe. A day is counted as **operational** if at least one in-range image exists (i.e. `image_fov != "OOR"`). Images triggered by `"CodeLoc Not Entered"` are excluded entirely.

```{r op-days}
df_days <- df_image_report |>
  cam_get_op_days(
    grouping   = c("project_id", "project", "location_id", "location"),
    missing_as = TRUE,
    span       = "data"
  )
```

The result is a long tibble with one row per camera per day, with an `operating`
logical column.

```{r op-days-ex, scroll=TRUE}

head(df_days, 10)

```

### Summarising by season

`cam_summarise_op_by_season()` aggregates the daily calendar built in the previous step into seasonal operating-day totals. Seasons are defined by **Julian day start cutoffs**. For example, the standard ABMI season definitions are the Julian days of 99, 143, and 288, which correspond to the beginning of the spring, summer, and winter periods for downstream EDD calculations. 

```{r op-season}
# Standard ABMI season definitions
seasons <- c(spring = 99L,   # ~April 9
             summer = 143L,  # ~May 23
             winter = 288L)  # ~October 15

df_days_summary <- df_days |>
  cam_summarise_op_by_season(
    seasons  = seasons,
    by_year  = FALSE,
    wide     = TRUE   # one column per season + total_days
  )
```

The wide output has one row per camera location with columns `spring`, `summer`,
`winter`, and `total_days`, which are all counts of operational days.

```{r op-season-ex, scroll=TRUE}

head(df_days_summary, 10)

```

---

## Step 3: Consolidate species tags

WildTrax stores tags at the **image × individual tag** level: if two deer are
visible in one image and have different tag categories (e.g., one adult and one juvenile), there will be two rows in the main report. `cam_consolidate_tags()` collapses these into one row per **image × species**, summing counts and concatenating
age/sex classes, which is what is needed for downstream calculations.

```{r consolidate}
df_main_report_cons <- cam_consolidate_tags(df_main_report)
```

::: {.callout-note}
The function also detects and warns about likely duplicate tag entries (identical
rows except `tag_id`) and drops unknown `species_common_name` labels with a
warning — both indicate upstream data quality issues worth investigating.
:::

---

## Step 4: Time in front of camera

### Parse into series

`cam_calc_time_by_series()` groups consecutive images of the same species at the same camera into **series** using a time-gap threshold. Images more than `split_gap_secs` seconds apart (default 120 s) start a new series. Each series is then assigned a duration using the time-in-front-of-camera (TIFC) method, with several adjustments to account for probabilities that the animal left the camera field of view:

- **NONE-bridged gaps**: `cam_obtain_n_gap_class()` is called internally to detect cases where an animal briefly leaves the field of view — and a `NONE` image was triggered — and returns within the gap threshold. Without this check, such a pair would be collapsed into a single series and over-estimate time spent. Any detected boundary forces a new series regardless of the time gap.
- **Probabilistic gap adjustment**: within-series gaps of 20–120 s fall in an ambiguous zone — too short to confidently split into a new series, but long enough that the animal may have temporarily left the field of view. By default (`adjust_gap_prob = TRUE`), a species-group-specific probability of departure is applied to both sides of these gaps, reducing the time credited across them proportionally. Set `adjust_gap_prob = FALSE` to disable this and compare results under both assumptions.

```{r series}
df_series <- cam_calc_time_by_series(df_main_report_cons)
```

The output is one row per series with columns `project`, `location`,
`species_common_name`, `series_num`, `n_images`, `series_total_time` (seconds), `series_start`, and `series_end`.

```{r series-ex, scroll=TRUE}

head(df_series, 10)

```

### Sum to seasonal totals

`cam_sum_total_time()` rolls up series-level times to seasonal totals per camera × species, zero-filling locations with no detections for any species in the`species_universe` argument.

```{r total-time}
sp_uni <- c("White-tailed Deer", "Black Bear", "Moose",
            "Woodland Caribou", "Fisher", "Marten",
            "Coyote", "Snowshoe Hare", "Canada Lynx")

df_dur <- df_series |>
  cam_sum_total_time(
    season_cutoffs   = seasons,
    op_days_df       = df_days_summary,
    species_universe = sp_uni
  )
```

The output has one row per camera × species × season with `total_duration` (seconds of TIFC) and `total_season_days` (operational days in that season).

```{r total-time-ex, scroll=TRUE}

head(df_dur, 10)

```

---

## Step 5: Calculate density

### Camera model lookup

EDD values vary by camera model. `cam_extract_model_lookup()` derives a `model` label (`"hf2"` or `"pc900"`) for each camera from the `equipment_model`field in the image report.

```{r model}
df_model <- df_image_report |>
  cam_extract_model_lookup(
    keys      = c("project", "project_id", "location", "location_id"),
    model_col = "equipment_model"
  )
```

### Density calculation

`cam_calc_density_by_loc()` joins EDD values from the internal lookup table to each camera × species × season row and applies the TIFC density formula. EDD lookups are keyed on species distance group, vegetation category (`overall_category`), season, camera model, and mounting height. Users can adjust the camera angle height, the format of the output (long or wide), and choose whether or not to aggregate the density value at each location across the seasons sampling. The output dataframe reports density at each location as the number of individuals per km². 

Note that users must supply their own lookup table for each location's EDD vegetation category (i.e., `edd_category_df`), as well as indicate the height of the camera. 

```{r edd-categories, echo=FALSE}

library(googlesheets4)
library(googledrive)

sheet_id <- drive_find(type = "spreadsheet", 
                       shared_drive = "ABMI Mammals") |>
  filter(str_detect(name, "BDT")) |>
  select(id) |>
  pull()

df_bdt_edd <- read_sheet(ss = sheet_id) |>
  mutate(overall_category = paste0(primary_category, "_", 
                                   secondary_category)) |>
  select(-c(primary_category, secondary_category))

```

```{r density}
df_density <- df_dur |>
  left_join(df_model, by = c("project", "location")) |>
  mutate(height = "high") |>              # 1 m mounting height
  cam_calc_density_by_loc(
    edd_category_df     = df_bdt_edd,     # vegetation category per location
    cam_fov_angle       = 40,             # degrees
    format              = "long",
    aggregate           = TRUE,           # weighted mean across seasons
    use_global_edd      = TRUE,           # fall back to pooled EDD if needed
    annotate_edd_source = TRUE            # label EDD provenance
  )

head(df_density, 10)

```

#### Seasonal exclusions

By default, Black Bear detections in winter are excluded from the seasonal
weighted mean before aggregating (bears are hibernating and zero detections
do not reflect low density). This is controlled by `agg_exclude_species` and
`agg_exclude_season`:

```{r exclude, eval=FALSE}
# Default: exclude Bear × winter
cam_calc_density_by_loc(
  ...,
  agg_exclude_species = "Bear",
  agg_exclude_season  = "winter"
)

# To include all seasons:
cam_calc_density_by_loc(
  ...,
  agg_exclude_species = NULL
)
```

---

## Full pipeline at a glance

```{r full-pipeline, eval=FALSE}
# 1. Operating days
df_days <- df_image_reports |>
  cam_get_op_days(grouping = c("project_id", "project", "location_id", "location"),
                  span = "data", missing_as = TRUE)

df_days_summary <- df_days |>
  cam_summarise_op_by_season(seasons = c(spring = 99L, summer = 143L, winter = 288L),
                             wide = TRUE)

# 2. Consolidate tags
df_main_reports_cons <- cam_consolidate_tags(df_main_reports)

# 3. Time in front of camera
df_series <- df_main_reports_cons |>
  cam_calc_time_by_series(split_gap_secs = 120)

df_dur <- df_series |>
  cam_sum_total_time(season_cutoffs   = c(spring = 99L, summer = 143L, winter = 288L),
                     op_days_df       = df_days_summary,
                     species_universe = sp_uni)

# 4. Density
df_model <- df_image_reports |>
  cam_extract_model_lookup(keys = c("project", "location"))

df_density <- df_dur |>
  left_join(df_model, by = c("project", "location")) |>
  mutate(height = "high") |>
  cam_calc_density_by_loc(edd_category_df     = df_bdt_edd,
                          cam_fov_angle       = 40,
                          aggregate           = TRUE,
                          use_global_edd      = TRUE,
                          annotate_edd_source = TRUE)
```
